---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# depmicrosimr

<!-- badges: start -->
<!-- badges: end -->

Dynamic electricity pricing (DEP) is a policy and regulatory tool to increase flexibility in power systems. The objective is to produce a near real-time system balancing effect. This enhances peak-shaving, lowers renewables curtailment, reduces storage requirements etc. DEP is a win-win because it has the potential to produce financial gains at both system and the individual household level. *depmicrosimr* is an agent-based model designed to (1) project the uptake of dynamic electricity tariffs on Irish consumers and (2) investigate consequences for the power system.

**Documentation:** <https://phalacrocorax-gaimardi.github.io/depmicrosimr/>

## Installation

You can install the development version of *depmicrosimr* using:

``` r
remotes::install_github("phalacrocorax-gaimardi/depmicrosimr")
```

## Example

*depmicrosimr* simulates load-shifting arising from a switch from flat to dynamic pricing using a penalised quadratic model and the OSQP solver. The key parameters are a behavioural cost penalty factor $\gamma$, a load-shift timescale $\tau$, and an inflexible load fraction $\phi$. A household can tolerate departures from the "natural" household load profile (taken to be their flat tariff profile) over periods shorter than $\tau$ with little penalty, but the penalty grows steeply over periods longer than $\tau$.

The example below simulates one year of dynamic prices and finds the load profile that minimises the bill of a household whose natural load is the standard urban profile LP1.

```{r example, results = 'hide', message = FALSE, warning = FALSE, fig.width = 8, fig.height = 2.5}
library(depmicrosimr)
library(dplyr)
library(ggplot2)

set.seed(1)
prices <- set_prices(sD)

# dynamic-tariff prices for 2026 with a 4,200 kWh/year household on profile LP1
demand <- prices %>%
  filter(tariff_plan == "dynamic", lubridate::year(datetime) == 2026) %>%
  inner_join(load_profiles_generalised %>% select(datetime, lp1)) %>%
  mutate(load = 4200 * lp1) %>%
  select(datetime, price, load)

# 50% of load is flexible, behavioural cost parameter 0.5, loads shiftable over about 24 h
demand_response <- get_flex(demand, phi = 0.5, tau = 24, gamma = 0.5)

# plot the unshifted and shifted loads for February
demand_response %>%
  filter(lubridate::yday(datetime) %in% 32:60) %>%
  ggplot() +
  geom_line(aes(datetime, load_opt), colour = "orange") +
  geom_line(aes(datetime, load), colour = "grey50", linetype = "dotted")
```

See the [Getting started](https://phalacrocorax-gaimardi.github.io/depmicrosimr/articles/getting-started.html) article to run the full agent-based model, and the [Model description](https://phalacrocorax-gaimardi.github.io/depmicrosimr/articles/model-description.html) for how it works.
