---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# depmicrosimr

<!-- badges: start -->
<!-- badges: end -->

Dynamic electricity pricing (DEP) is a policy and regulatory tool to increase flexibility in power systems. The objective is to produce a near real-time system balancing effect. This enhances peak-shaving, lowers renewables curtailment, reduces storage requirements etc. DEP is a win-win because it has the potential to produce financial gains at both system and the individual household level. *depmicrosimr* is an agent-based model designed to (1) project the uptake of dynamic electricity tariffs on Irish consumers and (2) investigate consequences for the power system.

## Installation

You can install the development version of *depmicrosimr* using:

``` r
remotes::install_github("phalacrocorax-gaimardi/depmicrosimr")
```

## Examples

*depmicrosimr* simulates load-shifting arising from a switch from flat to dynamic pricing using a penalised quadratic model and the OSQP solver. The key parameters are a behavioural cost penalty factor $\gamma$, a load-shift timescale $\tau$, and an inflexible load fraction $\phi$. A household can tolerate departures from the "natural" household load profile (taken to be their flat tariff profile) over periods shorter than $\tau$ with little penalty, but the penalty grows steeply over periods longer than $\tau$.

Wholesale prices are contained in *sem_prices_2023_2025*. The example below assumes the standard urban load profile LP1 from the dataset *load_profiles*.

```{r example,results='hide', message=FALSE, warning=FALSE,fig.width=8, fig.height=2}
library(depmicrosimr)
#optimise load-shifting behaviour based on 2025 wholesale prices for a household with "natural" load profile LP1
demand <- make_demand_response_data(profile="lp1",mean_daily_load=20,years=2025)
#assume 50% of load is flexible, behavioural cost parameter is 0.5, loads are shiftable over 24h period.
demand_response <- get_flex(demand,phi=0.5,tau=24,gamma=0.5)
#add a day of year column
demand_response <- demand_response %>% dplyr::mutate(yday=lubridate::yday(datetime))
#plot the unshifted and shifted loads for February
library(ggplot2)
dr <- demand_response %>% dplyr::filter(yday %in% 32:60) 
dr %>% ggplot()+geom_line(aes(datetime,load_opt),colour="orange")+geom_line(aes(datetime,load),colour="grey50",linetype="dotted")

```


```{r plot, echo = FALSE}

```

