---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The examples below simulate prices and solve many optimisation problems, so they are
# slow. They are shown but not run when the site is built. To run them yourself, set
#   Sys.setenv(DEPMICROSIMR_RUN_VIGNETTES = "true")
# before building, or copy the code into an R session.
run_heavy <- identical(Sys.getenv("DEPMICROSIMR_RUN_VIGNETTES"), "true")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = run_heavy,
                      fig.width = 7, fig.height = 3.2)
```

This article shows the three things most users do first: simulate a price scenario,
see how one household responds to it, and run the full agent-based model. For how
the model works, see `vignette("model-description")`.

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("phalacrocorax-gaimardi/depmicrosimr")
```

Load the package. Its scenario table `sD` is the base scenario used throughout.

```{r load, eval = TRUE}
library(depmicrosimr)
head(sD[, c("parameter", "value", "unit")])
```

## 1. Simulate retail prices

`set_prices()` simulates one wholesale price path from 2026 onwards (observed prices
before that) and turns it into hourly **flat**, **ToU** and **dynamic** tariffs.
It draws random numbers, so every call gives a different path.

```{r prices}
set.seed(1)
prices_scen <- set_prices(sD)
head(prices_scen)
```

## 2. How one household responds

`get_flex()` takes a table with columns `datetime`, `price` and `load` (the household's
natural load, in kWh per hour) and returns the optimal shifted load `load_opt`.
The parameters are the inflexible fraction `phi`, the behavioural cost weights `gamma` and `eta`, and the load-shift timescale
`tau` in hours.

```{r get-flex}
library(dplyr)

demand <- prices_scen %>%
  filter(tariff_plan == "dynamic", lubridate::year(datetime) == 2026) %>%
  inner_join(load_profiles_generalised %>% select(datetime, lp1)) %>%
  mutate(load = 4200 * lp1) %>%          # a household using 4,200 kWh a year
  select(datetime, price, load)

response <- get_flex(demand, phi = 0.4, gamma = 5, eta = 0.3, tau = 24)

# share of annual energy that was moved to another hour
100 * sum(abs(response$load_opt - response$load)) / (2 * sum(response$load))
```

Plot one winter week of natural against shifted load:

```{r plot-flex}
library(ggplot2)
week <- response %>% filter(lubridate::yday(datetime) %in% 32:38)

ggplot(week, aes(datetime)) +
  geom_line(aes(y = load, colour = "natural")) +
  geom_line(aes(y = load_opt, colour = "shifted")) +
  scale_colour_manual(values = c(natural = "grey50", shifted = "orange")) +
  labs(x = NULL, y = "kWh per hour", colour = NULL)
```

Larger `gamma` makes shifting costlier, so load moves less. Larger `tau` lets the
household move load further in time before it is penalised. Increasing `phi` makes more
of the load inflexible.

## 3. Run the agent-based model

`runABM()` runs one or more independent simulations. Each builds its own price path,
social network and agent population, then steps forward in two-month intervals from
January 2019. Dynamic tariffs are introduced in mid-2026, so the run has to extend past
that date to see any dynamic adoption.

```{r run-abm}
set.seed(42)
res <- runABM(
  sD,
  Nrun            = 2,
  simulation_end  = 2030,
  behavioural_model = "prospect",   # "full" also counts the behavioural penalties
  use_parallel    = FALSE           # TRUE uses several cores (macOS/Linux/Windows)
)

names(res)          # "abm" "scenario" "system"
dim(res$abm)
```

Start with `Nrun = 1` and a short horizon: every sampled household solves several
optimisation problems at each step.

`res$abm` has one row per household, per date and per simulation. To see adoption over
time, count households by tariff plan:

```{r adoption}
adoption <- res$abm %>%
  count(simulation, date, tariff_plan) %>%
  group_by(simulation, date) %>%
  mutate(share = n / sum(n)) %>%
  ungroup()

ggplot(adoption, aes(date, share, colour = tariff_plan, linetype = factor(simulation))) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "share of households", colour = "tariff plan", linetype = "run")
```

Before smart meters are installed, households are on `flat` or the old `tou_old`
day/night tariff. `tou` appears as smart meters are installed, and `dynamic` after
mid-2026.

## Things to know

* **Reproducibility.** The model is stochastic at every level. `set.seed()` fixes a
  single-process run (`use_parallel = FALSE`). Parallel runs are not reproducible with
  R's default random number generator.
* **Price paths are not returned.** `runABM()` returns the household states but not the
  price path that each run used.
* **Choosing a behavioural model.** `"prospect"` and `"full"` are the current models.
  `"full"` is the default in `runABM()`; see `vignette("model-description")`, section 7.
* **Scenarios.** `sD` is the base scenario. Other scenarios are sheets in
  `inst/ext_data/scenario_parameters.xlsx` and can be read with `readxl::read_xlsx()`
  into a data frame with the same columns as `sD`.
