---
title: "Model description"
subtitle: "How depmicrosimr simulates the uptake of dynamic electricity pricing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model description}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 3.6)
library(depmicrosimr)
```

This article describes the model **as implemented in the code**. Function names
link to the reference pages. Parameter values quoted here are the defaults in the
`Base` sheet of `inst/ext_data/scenario_parameters.xlsx`, available in the package as
the dataset `sD`.

## 1. What the model does

*depmicrosimr* is an agent-based model (ABM) of Irish households choosing between
electricity tariffs. It has two aims: to project how many households move from a
**flat** tariff to a **time-of-use (ToU)** or **dynamic** tariff, and to describe the
**load shifting** that follows, so that the consequences for the power system can be
studied.

The model combines five ingredients:

| Ingredient | In one line | Main functions |
|---|---|---|
| Agents | About 1,200 synthetic households built from a 2024 household survey | `initialise_agents()` |
| Prices | Simulated wholesale prices and a retail model that turns them into flat, ToU and dynamic tariffs | `sem_prices()`, `set_prices()` |
| Load shifting | A household shifts load to cut its bill, subject to behavioural costs | `get_flex()` |
| Decision rule | Households compare plans using prospect theory and are influenced by their neighbours | `evaluate_tariffs()`, `evaluate_full_cost()`, `get_ce_value()` |
| Social network | A homophilous network over which tariff choices spread | `make_artificial_society()` |

## 2. The simulation loop

Time runs in **bi-monthly steps (six per year) from January 2019**. 2019 is before
smart-meter rollout, so households start on a flat tariff or on the old day/night
tariff. Dynamic pricing is introduced in mid-2026 (`dep_introduction = 2026.5`).

```
runABM(scen, Nrun, simulation_end, ...)
 └─ for each simulation j = 1..Nrun            (parallelisable, independent)
     ├─ prices_scen  <- set_prices(scen)                 # new price path
     ├─ social       <- make_artificial_society(...)     # new network
     ├─ agents_0     <- initialise_agents(...)           # new agent draw
     └─ for t = 2..Nt                                    # one step = 2 months
         └─ agents_t <- update_agents(agents_{t-1}, ...)
```

At each step `update_agents()`:

1. builds the scenario parameters for that date (`scenario_params()`);
2. switches households on the old day/night tariff (`tou_old`) to `tou` once
   their smart meter has been installed (`yeartime >= rollout`);
3. draws a random sample of households, a fraction `p.` of the population
   (`p. = 0.03`, an inertia parameter described in the scenario sheet as estimated from CRU
   switching data; stochastic rounding is used, see `roundr()`). Only these
   households re-evaluate their tariff in this step;
4. for each sampled household, evaluates the available plans and picks one
   (section 6);
5. recomputes each household's count of neighbours on ToU and dynamic tariffs.

Each run therefore uses its own price path, its own network and its own random agent
attributes, so variation between runs reflects all three sources of uncertainty.

## 3. Agents

### 3.1 Where agents come from

Agents are the respondents of a 2024 household survey (Li et al., 2024), stored as
`dep_survey`. `initialise_agents()` builds the starting state as follows.

* **Annual consumption** (`kWh`) is inferred from each respondent's reported
  highest and lowest bi-monthly electricity bills, converted to kWh with 2023 SEAI
  prices, and fitted with a sinusoidal seasonal demand model
  (`survey_bills_to_kwh()`, `demand_fun()`). Missing bills are imputed
  stochastically, so `kWh` varies from run to run.
* **Initial tariff.** Respondents who report "don't know", "other" or a day/night/peak
  tariff are treated as `flat`. Of the respondents on a day/night tariff, two thirds
  are assumed to be on the old day/night tariff (`tou_old`) in 2019 and the rest on
  `flat`. Prepay customers are removed.
* **Area and rollout.** Respondents are classed as urban or rural and linked to a
  county, which gives an estimated smart-meter rollout date (`smart_meter_rollout`).
* **Natural load profile.** Urban households have profile `lp1`, rural households `lp3`.
  Households on the old day/night tariff have the measured ToU profiles `lp2`
  (urban) and `lp4` (rural). See `load_profiles_generalised`.
* **Latent traits.** Three traits (inertia, proactiveness, perceived flexibility)
  come from a structural equation model of the survey (Kumar et al., 2026), stored in
  `struct_params`.
* **Awareness.** Each household becomes aware that dynamic tariffs exist at a time
  drawn uniformly between 2026.5 and `dyn_aware_year` (2028.5 in the base scenario).
  Before that date the dynamic plan is not in its choice set.

### 3.2 Behavioural parameters

Each household gets the following, using the traits above.

* **Flexibility parameters** $(\gamma, \tau)$. The survey-derived flexibility trait
  is standardised and mapped through a Beta distribution to a *one-hour flexibility
  score* between zero and the largest score attainable for the chosen $(\phi, \eta)$.
  The score is then matched to a $(\gamma,\tau)$ pair from a pre-computed table
  (`flex_scores`, `flex_score_cube()`, `match_flex_params()`). The mapping is
  one-to-many, so the match is randomised among the two nearest grid points. $\phi$
  and $\eta$ are the same for all households (defaults 0.4 and 0.3).
* **Loss aversion** $\lambda$. The inertia trait is mapped through a Beta
  distribution to the interval from `pt_min_loss_aversion` (1.25) to
  `pt_max_loss_aversion` (4.25).
* **Social variables.** `degree` is the household's number of network contacts, and
  `q_tou`, `q_dyn` are how many of them are on a ToU or dynamic tariff.

The calibration target for aggregate flexibility is the difference between the
standard flat (`lp1`) and ToU (`lp2`) load profiles (14.4%). `initialise_agents()` prints
summary statistics of the household flexibility scores next to that target.

## 4. Prices

### 4.1 Wholesale prices

`sem_prices()` returns hourly wholesale prices for the Single Electricity Market:
observed prices for 2019 to 2025, then a simulation. Prices are modelled on a
**log-like scale** that tolerates negative prices,

$$
y_t = \operatorname{arcsinh}\!\left(\frac{p_t}{s}\right), \qquad s = 0.01~\text{€/kWh},
$$

so that $p_t = s \sinh(y_t)$. The simulated series is a sum of four parts:

$$
y_t = \underbrace{\text{trend}_t}_{\text{scenario}} + \underbrace{\text{day}_t + \text{week}_t + \text{year}_t}_{\text{STL seasonality, 2019--2025}} + \underbrace{\varepsilon_t}_{\text{3-state Gaussian HMM}} .
$$

* The **trend** is anchored at the end-2025 trend and at scenario values for the
  wholesale price at end-2030 (0.13 €/kWh) and end-2040 (0.20 €/kWh), interpolated
  linearly on the $y$ scale. It is flat after 2040.
* The **seasonal terms** (daily, weekly, annual) are taken from an STL decomposition
  of the 2019–2025 data (`decompose_logprices()`, stored in `sem_logprices_2019_2025_decomp`).
* The **noise** $\varepsilon_t$ is drawn from a pre-fitted three-state Gaussian hidden
  Markov model of the decomposition residuals (`hmm_fit`, `generate_logprice_hmm()`,
  `simulate_hmm()`). The states are intended to capture high-renewables, normal and
  low-renewables regimes. Because the HMM is Gaussian on the log-like scale, price
  variance is state-dependent.
* An optional **price shock** (`shock = TRUE`) multiplies the trend price by four
  from 1 January 2030, decaying with a one-year time constant.

The price path does not respond to household behaviour: there is no feedback from
aggregate load shifting to prices.

### 4.2 Retail tariffs

`set_prices()` turns wholesale prices into three consistent retail tariffs on an
hourly grid, for the whole simulated period:

* **Dynamic.** Wholesale price, capped, plus a time-of-day network charge, plus VAT:

$$
p^{\text{dyn}}_t = \bigl(1 + \text{VAT}_t\bigr)\Bigl[\min\bigl(\text{cap}_t,\; p^{\text{sem}}_t\bigr) + n_{b(t)}(t)\Bigr].
$$

  The cap starts at 0.5 €/kWh in 2026 (the CRU dynamic price cap) and scales with the
  wholesale trend price. $n_b(t)$ is the network charge for the hour's ToU band
  $b \in \{\text{night},\text{day},\text{peak}\}$, interpolated between scenario values
  for 2020, 2026, 2030 and 2040. The ToU bands are in the package data `load_profiles_generalised`
  (night 23:00–08:00, peak 17:00–19:00, day otherwise, adjusted for daylight saving).
* **Flat.** A single price that follows a *trailing 365-day* load-weighted mean of the
  dynamic price (with a 2% hedging premium), using a weighted mixture of the standard
  flat and ToU load profiles (`lp1`, `lp2`). The flat tariff therefore lags the market.
* **ToU.** The same trailing average, computed separately within each ToU band.

The parameter `w` (default 1/3 in `set_prices()` and `runABM()`) sets how much of the
value of flexibility suppliers retain, so that a household with no flexibility can
be worse off on ToU or dynamic than on flat. VAT is 13.5% before 2022.4, 9% from
2022.4, and 13.5% from 2031 (`vat_rate_fun()`). Standing charges (`standing_charge_fun()`)
rise from €130 (2015) to €350 (2040); the dynamic plan adds a premium (€75 in the base
scenario).

Each household's bill is price times load, summed over 12 months, plus the standing charge.

## 5. How a household shifts load

The core of the model is `get_flex()`. Given a price series and a natural load profile
for a household, it finds the hourly load that minimises **bill plus behavioural
cost**.

### 5.1 The optimisation problem

Over a horizon of $T$ hours (one year), let $L^0_t$ be the household's natural load
(its profile under flat prices) and $x_t$ the change from it, so that the shifted load
is $L_t = L^0_t + x_t$. Total annual energy is $E = \sum_t L^0_t$. The household solves

$$
\min_{x}\;\; \sum_t p_t\,x_t \;+\; \frac{\tilde\eta}{2}\sum_t (x_{t+1}-x_t)^2 \;+\; \frac{\tilde\gamma}{2}\sum_t \Bigl(\sum_k g(k-t)\,x_k\Bigr)^{2}
$$

subject to

$$
\sum_t x_t = 0, \qquad -(1-\phi_t)\,L^0_t \;\le\; x_t \;\le\; P_{\max} - L^0_t .
$$

In words:

* **Bill term** $\sum_t p_t x_t$. Moving load to cheap hours lowers the bill.
* **Ramping term** (parameter $\eta$). Penalises abrupt hour-to-hour changes in the shift,
  which keeps the shifted profile smooth and stops the household reacting to tiny
  price differences.
* **Stiffness term** (parameter $\gamma$, timescale $\tau$). The kernel $g$ decays on a
  timescale $\tau$ hours, so $\sum_k g(k-t)x_k$ is a local average of the shift. A load
  that is moved *and then returned within about $\tau$ hours* nets out and costs little;
  load moved for much longer than $\tau$ is penalised heavily. Households with large
  $\tau$ and small $\gamma$ are more flexible.
* **Energy neutrality** $\sum_t x_t = 0$. The household changes *when* it uses
  electricity, not *how much*. The constraint holds over the whole horizon, with
  the stiffness term discouraging long-range shifts.
* **Bounds.** The household cannot shift more than its flexible load
  $(1-\phi_t)L^0_t$ and cannot exceed its maximum import capacity $P_{\max}$
  (`mic = 12` in the base scenario).

The scaled coefficients are $\tilde\eta = \eta\cdot 8760/E$ and
$\tilde\gamma = \gamma\,(8760/E)/\lVert g\rVert^2$, so that $\gamma$ and $\eta$ are
expressed for a household with a standard 8,760 kWh annual load. The kernel is
exponential by default, $g(\ell) = e^{-|\ell|/\tau}$, with Gaussian, Cauchy and Matérn
alternatives selectable through the `kernel` argument. The problem is a sparse convex
quadratic programme solved with OSQP (Stellato et al., 2020).

### 5.2 Inflexible load

A fraction $\phi$ of load is inflexible. It varies through the day with a factor $f_h$
(`diurnal_inflex`, estimated from the lowest quantile of smart-meter profiles of
households without solar PV, normalised to a mean of one). The inflexible share in hour
$t$ is

$$
\phi_t = \phi - (1-\phi)\ln(1-\phi)\,(f_{h(t)} - 1),
$$

which averages to $\phi$ and tends to one as $\phi \to 1$ whatever $f$ is.

### 5.3 Kernels and the shape of the penalty

```{r kernels, echo = FALSE}
tau <- 24
lag <- 0:96
plot(lag, exp(-lag / tau), type = "l", lwd = 2, xlab = "lag (hours)", ylab = "g(lag)",
     main = expression(paste("Exponential kernel, ", tau, " = 24 h")), ylim = c(0, 1))
lines(lag, exp(-(lag / (1.128 * tau))^2), lty = 2, lwd = 2)
lines(lag, 1 / (1 + (lag / (0.6366 * tau))^2), lty = 3, lwd = 2)
legend("topright", legend = c("exp (default)", "gauss", "cauchy"), lty = 1:3, lwd = 2, bty = "n")
```

### 5.4 Flexibility score

The package summarises how much a household shifts using a **flexibility score**, the
share of annual energy that is moved:

$$
\text{flex} = 100 \times \frac{\tfrac12 \sum_t |x_t|}{E} \;\%.
$$

`get_flex_scores()` computes it at one-hour resolution and for coarser windows, and the
precomputed table `flex_scores` links $(\phi,\eta,\gamma,\tau)$ to scores.

### 5.5 Behavioural cost

When the household's decision depends on the behavioural cost (behavioural model
`"full"`, section 7), the cost of the flexible plan is the optimised bill plus the two
penalty terms evaluated at the optimum:

$$
\text{cost}^{\text{flex}} = \sum_t p_t L_t + \text{standing charge}
 + \tfrac12\tilde\gamma \lVert Kx\rVert^2 + \tfrac12\tilde\eta \sum_t (x_{t+1}-x_t)^2 ,
$$

computed in `get_full_annual_cost()`.

## 6. Choosing a tariff

### 6.1 Which plans are available

* `flat`: always.
* `tou`: once the household's smart meter is installed (`yeartime >= rollout`).
* `dynamic`: once the household is aware of it (`yeartime >= dyn_aware`).

Households evaluate each plan on the **simulated hourly prices for the twelve months
starting at the decision date**. In effect they have perfect foresight over a one-year
horizon.

### 6.2 Costs under each plan

For ToU and dynamic plans the household computes two annual costs, relative to its flat bill
$B^{\text{flat}}$:

* $B^{\text{flex}}$, achieved if it **does** manage to shift load optimally, and
* $B^{\text{noflex}}$, if it **does not** shift and simply pays the new prices for its natural load.

The savings are $s^{\text{flex}} = B^{\text{flat}} - B^{\text{flex}}$ and
$s^{\text{noflex}} = B^{\text{flat}} - B^{\text{noflex}}$. The second can be a loss.

### 6.3 Prospect theory

Switching is treated as a risky prospect with two outcomes. The household succeeds in
shifting with probability $c$ (its *confidence*) and fails with probability $1-c$.
Outcomes are valued with the Tversky–Kahneman (1992) value function and probabilities
are distorted by a weighting function:

$$
v(s) = \begin{cases} s^{\alpha} & s \ge 0 \\ -\lambda\,(-s)^{\beta} & s < 0 \end{cases},
\qquad
w(c;\delta) = \frac{c^{\delta}}{\bigl(c^{\delta} + (1-c)^{\delta}\bigr)^{1/\delta}} .
$$

The prospect value and its euro-denominated **certainty equivalent** are

$$
\text{PV} = w(c;\delta^{+})\,v(s^{\text{flex}}) + w(1-c;\delta^{-})\,v(s^{\text{noflex}}),
\qquad
\text{CE} = \begin{cases} \text{PV}^{1/\alpha} & \text{PV}\ge 0 \\ -\bigl(-\text{PV}/\lambda\bigr)^{1/\beta} & \text{PV}<0 \end{cases}
$$

with $\alpha=\beta=0.88$, $\delta^{+}=0.61$ (used for the "flexibility achieved"
outcome) and $\delta^{-}=0.69$ (used for the "not achieved" outcome). These come from
`sD`. The loss-aversion coefficient $\lambda$ is **agent-specific** (section 3.2); the
value 2.25 in `sD` is only the default for stand-alone calls to `get_ce_value()`.

```{r pt-curves, echo = FALSE, fig.width = 8, fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
alpha <- 0.88; beta <- 0.88
v <- function(s, lambda) ifelse(s >= 0, abs(s)^alpha, -lambda * abs(s)^beta)
s <- seq(-100, 100, length.out = 401)
plot(s, v(s, 2.25), type = "l", lwd = 2, xlab = "saving vs flat tariff (EUR)", ylab = "value v(s)",
     main = "Value function")
lines(s, v(s, 1.25), lty = 2, lwd = 2); lines(s, v(s, 4.25), lty = 3, lwd = 2)
abline(h = 0, v = 0, col = "grey70")
legend("topleft", legend = c(expression(lambda == 2.25), expression(lambda == 1.25),
                             expression(lambda == 4.25)), lty = c(1, 2, 3), lwd = 2, bty = "n")
wf <- function(p, d) p^d / (p^d + (1 - p)^d)^(1 / d)
p <- seq(0, 1, length.out = 201)
plot(p, wf(p, 0.61), type = "l", lwd = 2, xlab = "probability", ylab = "weight w(p)",
     main = "Probability weighting")
lines(p, wf(p, 0.69), lty = 2, lwd = 2); abline(0, 1, col = "grey70")
legend("topleft", legend = c(expression(delta == 0.61), expression(delta == 0.69)),
       lty = 1:2, lwd = 2, bty = "n")
par(op)
```

### 6.4 Confidence and social influence

The confidence $c$ that a household will manage to shift load is not fixed. It rises
with the share of the household's network contacts who have already adopted:

$$
c^{\text{tou}} = c_{\text{tou}} + \min\!\Bigl(1, \tfrac{q_{\text{tou}}+q_{\text{dyn}}}{\text{degree}}\Bigr)\,(c_{\text{tou}}^{\max}-c_{\text{tou}}),
$$

$$
c^{\text{dyn}} = c_{\text{dyn}} + \min\!\Bigl(1, \tfrac{q_{\text{dyn}}}{\text{degree}}\Bigr)\,(c_{\text{tou}}-c_{\text{dyn}}),
$$

with base-scenario values $c_{\text{tou}} = 0.70$, $c^{\max}_{\text{tou}} = 0.85$ and
$c_{\text{dyn}} = 0.50$. A household whose neighbours all use ToU or dynamic tariffs is
therefore as confident about ToU as 0.85, and one whose neighbours all use dynamic
tariffs is as confident about dynamic as it would be about ToU without any social
reinforcement (0.70). Households with no contacts feel no social effect.

### 6.5 Decision rule

The household computes $\text{CE}^{\text{tou}}$ and $\text{CE}^{\text{dyn}}$ for the plans
open to it. It adopts the plan with the larger certainty equivalent **if that is
positive**; otherwise it chooses `flat`. A sampled household can therefore also move
*back* to flat.

## 7. Behavioural model variants

The `behavioural_model` argument of `runABM()` and `update_agents()` selects how costs
enter the decision:

| Value | Flexible-plan cost | Status |
|---|---|---|
| `"prospect"` | Optimised bill only (`evaluate_tariffs()`) | Current |
| `"full"` | Optimised bill **plus** the $\gamma$ and $\eta$ behavioural penalties (`evaluate_full_cost()`) | Current; the default in `runABM()` |
| `"classic"` | Cheapest-plan rule with a social term and a threshold $\theta$ | Legacy |

## 8. Social network

`make_artificial_society()` builds a random **homophilous** network over the agents
(Gower distance on the demographic variables in `dep_society_1`, weighted by `homophily`).
The probability that two agents are linked falls with their social distance $d_{ij}$:

$$
P(i \sim j) \propto (1-d_{ij})^{\nu}, \qquad \nu = 4.5,
$$

rescaled so that each agent's expected number of contacts approximately matches the
degree reported in the survey. Agents who reported no social contacts stay isolated. The network is
**static** within a run. `get_network_characteristics()` tabulates its assortativity
by demographic variable.

## 9. Scenarios

The scenario workbook has one sheet per scenario. `sD` is the `Base` sheet.

```{r sD-table, echo = FALSE}
knitr::kable(sD[, c("category", "parameter", "value", "unit")], digits = 5,
             caption = "Parameters in the base scenario `sD`.")
```

The other sheets in the workbook are variations on the base. According to the workbook's
`Description` sheet, `EDSC` and `LDSC` vary the dynamic standing-charge premium,
`LowPrice` and `HighPrice` change the wholesale price path for 2030 and 2040, and
`NoToU` and `Battery` are further intended variants. Check the sheet values against the
descriptions before relying on a scenario; the workbook is
`inst/ext_data/scenario_parameters.xlsx` in the repository.

## 10. Outputs

`runABM()` returns a list of three elements.

* `abm`: a data frame with one row per agent, per time step and per simulation. It holds the
  household's state at that date: `serial`, `kWh`, `tariff_plan`, `rollout`, `area`,
  `phi`, `eta`, `gamma`, `tau`, `lambda`, `flex_score`, `natural_profile`, `profile`, `annual_bill`,
  `degree`, `q_tou`, `q_dyn`, `dyn_aware`, the confidences `c_tou_soc` and `c_det_soc`,
  the certainty equivalents `CE_tou` and `CE_det` where computed, `simulation`, and `date`.
  The exact set of columns depends on the behavioural model.
* `scenario`: the scenario table that was used.
* `system`: the run settings (number of runs, end year, inertia `p.`, behavioural model,
  shock, `w`, and the confidence parameters).

Adoption curves are obtained by counting `tariff_plan` by `date`. See
`vignette("getting-started")`.

## 11. Assumptions and limitations

These follow directly from the code and are worth keeping in mind when interpreting results.

* **Household flexibility is a modelling construct.** $(\gamma,\tau)$ are matched to stated
  survey flexibility, not estimated from observed load-shifting behaviour. Many parameter pairs
  give similar flexibility scores.
* **One-year perfect foresight.** Households see the simulated price path of the next
  twelve months when they decide.
* **Annual energy neutrality.** Load can in principle move across months, discouraged only
  by the stiffness penalty.
* **Two standard profiles.** Households follow an urban (`lp1`) or rural (`lp3`) natural
  profile scaled by their annual consumption. There is no explicit representation of solar PV,
  batteries, electric vehicles or heat pumps in the code.
* **Exogenous prices.** Wholesale prices are not affected by adoption.
* **Static network.** Links do not form or break.
* **Random inertia.** A household reconsiders its tariff on average once every
  $1/p. \approx 33$ steps, about five and a half years.
* **Results are unweighted.** `initialise_agents()` does not use the survey weights
  (`dep_survey$weights`).

## References

* Li et al. (2024). Household survey on dynamic electricity pricing in Ireland. *[full reference to be added]*
* Kumar et al. (2026). Structural equation model of attitudes to dynamic pricing. *[full reference to be added]*
* Stellato, B., Banjac, G., Goulart, P., Bemporad, A. and Boyd, S. (2020). OSQP: an operator
  splitting solver for quadratic programs. *Mathematical Programming Computation* 12(4), 637–672.
* Tversky, A. and Kahneman, D. (1992). Advances in prospect theory: cumulative representation
  of uncertainty. *Journal of Risk and Uncertainty* 5, 297–323.
